Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-11-26 paper/revision2/ResponseToReviewers.tex

\noindent \textbf{Response letter for Submission JML-24-17, 2nd round of revisions}
\vskip 2em
\noindent Dear Prof. Jarrold, 
\vskip 1em
with my co-authors Shravan Vasishth and Himanshu Yadav, I would like to submit the second revision of our article titled ``Do syntactic and semantic similarity lead to interference effects? Evidence from self-paced reading and event-related potentials using German''.

We are grateful for the thoughtful and constructive comments and suggestions from yourself and the reviewers, and feel that the quality of the manuscript has been greatly improved by the revisions.
To address the concern that we overstate the implications of our findings, we have clarified at various places in the abstract and discussion of our results that our findings only apply to the present design.

Clarifying text has been added in this second round of revisions (e.g., regarding the implications, the prolonged reading times starting at the distractor and the use of syntactic and semantic cues), and we have shortened the manuscript length by four pages. We have removed redundant text passages (e.g., in the description of the SPR results).

The OSF repository has been made public to ensure general accessibility. We have updated the link in the manuscript accordingly.

@@

\begin{quote}
    \Paste{thisdesign2}
\end{quote}

\begin{quote}
    \Paste{thisdesign3}
\end{quote}

\subsection*{Comment E.3}
\begin{quote}
``\textbf{Reviewer 2's additional point 1 that asks whether you have changed that particular analysis}.''
\end{quote}

\subsubsection*{Response to E.3}
Yes, we changed all analyses in the first revision. As we had explained in the response to reviewers on August 29, 2024 (specifically in the response letter and our responses to the comments E.2 and R3.5), the priors were changed from directional priors in the original version to symmetrical priors in the revised version. This was necessary because we added trial id as a predictor as was suggested by a reviewer. There was no a priori hypothesis of the sign of the effect of trial id (do the participants get faster or slower over the course of the experiment?), therefore the prior for trial id needed to be symmetrical (positive and negative values). It is not possible to include symmetrical and directional priors within a single brms model \citep{brms}, thus all priors were changed to symmetrical ones.

Additionally, we changed the method with which the Bayes factors were computed. Initially, we had used bridge sampling, but in the revision, we switched to the Savage-Dickey method because only using that method we were able to compare models with random slopes (this was also explained in the previous response letter, Response to R1.7). These changes did not change the results qualitatively, but led to some numerical changes which probably piqued the reviewer's interest.

@@

\section*{Reviewer \#2} 
\subsection*{Comment R2.1}
\begin{quote}
``I appreciate the authors' responses to my previous comments. However, despite the revisions, significant weaknesses remain, rendering in my opinion this study primarily a methodological contribution that adds little value to our understanding of interference effects. Additionally, the authors' strong claims often come through as overstatements that could result in serious misconceptions.
My overall assessment is as follows.

1. Interpretability of the SPR results: The results from the self-paced reading (SPR) task remain unclear. The only observable effect—the semantic interference—emerges well before the critical region of interest (specifically, at the distractor region) and persists without change beyond the critical region. As a result, any effect at the critical region is uninterpretable.

The authors acknowledge the interpretability issue with the SPR results in their response to my comment, and they now include the following statement: ``Given the pre-critical reading time differences, effects in the later regions, i.e., the critical and spill-over regions, cannot be attributed clearly to the stimuli in these regions and sentence processing mechanisms associated with them. Therefore, we refrain from further discussing effects occurring in the later regions." \textbf{However, this comes after more than 10 full pages (pp. 25-37) of discussion on the reaction times (RTs) in the critical and pre-critical regions, which would lead any reader to believe these findings are meaningful. If, as the authors now realize, the effect emerges much earlier, this entire discussion is misleading: the SPR results after the distractor region are simply uninterpretable.}''
\end{quote}

\subsubsection*{Response to R2.1}
We agree with the reviewer that the presentation and discussion of the SPR results was misleading and have revised it accordingly. We have revised Figure 3 to remove the focus on the critical region (see below). At the beginning of the Results section, we now anticipate our interpretation of the results and state that we believe that the distractors led to encoding interference. We also state that the results at the critical (and spill-over) region are confounded because it is not possible to disentangle the long-lasting encoding effect from retrieval effects potentially arising in the critical region.

@@

\setcounter{figure}{2}

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C shows the reading times of the sentence focusing on the regions between the distractor and critical verb for all conditions.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled_zoom.jpg}
\end{sidewaysfigure}
\clearpage

The reviewer seems to believe that our results are not publication-worthy because encoding interference confounded the reading times of all regions after the distractor. However, as we had already discussed in the results of the SPR results, the results of the two published studies using the present design \citep{vandyke07, mertzen} show indications that they were affected by encoding interference as well (see page \pageref{previous_encoding}):

\begin{quote}
    \Paste{previous_encoding}
\end{quote}

Therefore, our study replicates the encoding interference effect of the previous studies but it is the only study to prominently address that the reading times of the critical region are confounded by encoding interference. To ensure comparability to the previous studies, we believe that it is mandatory that we report the reading times effects in the same regions which have been reported by them: the pre-critical, critical and spill-over region -- despite the confound discussed above. We have revised the Results section to be more concise, starting on page \pageref{revised_SPR}:

Because the finding that the self-paced reading times were primarily affected by semantic encoding interference is a central finding of our study, we think that it is appropriate that we investigated it. We did not analyze the distractor itself and the immediately following words because these were not identical across conditions. Instead, we used the reading times of the pre-critical region which was the same word across conditions (\textit{abends}, `in the evening', in the example item used for the x-axis labels in Figure 3 above) as a proxy to investigate the effect of encoding interference caused by the distractor. This pre-critical word preceded the critical verb, thus the reading times at the pre-critical word are unlikely to show retrieval interference. To emphasize this we have renamed the former section ``Pre-critical effects" to ``Reading times difference prior to retrieval". We have revised the beginning of the Discussion on page \pageref{prior_to_retrieval} accordingly:

@@

\begin{quote}
    \Paste{gendisccaveat2}
\end{quote}

 We feel that the above quote and the other re-statements in the paper (including the abstract) appropriately limit the claims we make in the paper. We never state anywhere in the paper that syntactic interference is generally absent in sentence processing.

\subsection*{Comment R2.4}
\begin{quote}
``3. Misalignment of SPR and ERP results: The results from the self-paced reading (SPR) and event-related potential (ERP) methods do not align, with the only observed effect—semantic interference—pointing in opposite directions: inhibitory in SPR but facilitatory in ERP. Specifically, the authors report a reversed effect in the P600 component, showing a reduced P600 in conditions of higher semantic interference. This contradicts typical predictions, where the P600 is usually greater under high interference. The authors interpret this as a facilitatory interference effect (i.e., easier processing with high semantic interference). However, if this were the case, \textbf{we would also expect faster reaction times (RTs) in high semantic interference conditions, while the opposite is observed in the SPR data}.''
\end{quote}

\subsubsection*{Response to R2.4}
The SPR results and the ERP results in the N400 window both showed inhibitory interference, i.e., more processing difficulty under high vs.\ low semantic interference, manifesting in longer reading times and a more negative N400. We explain below in our response to R2.5 that the reduced P600 for high interference might be explained by facilitatory interference because under high semantic interference whatever noun was retrieved can be easily integrated.

In general, SPR results cannot show the same complex results pattern that the ERP results showed because they provide just one measure per region. Additionally, only at the critical (or spill-over) region facilitated integration might be observed and in line with the reviewer's other comments, we do not interpret the reading times of the these regions because they were overshadowed by the reading times differences starting at the distractor.

@@

\subsubsection*{Response to R2.5}

We have revised our discussion of the ERP results on page \pageref{inhibition_facilitation} to clarify how the seemingly conflicting results can be reconciled (see below). Interference might enhance the N400 amplitude because it makes memory retrieval more difficult and it might reduce the P600 because no matter which noun was retrieved it can be easily integrated with the verb.

\begin{quote}
    \Paste{inhibition_facilitation}
\end{quote}

\subsection*{Comment R2.6}
\begin{quote}
``5. Unclear conflicting mechanisms: The authors attribute the SPR results to encoding interference and the ERP results to retrieval interference, leading to a contradictory conclusion. I won't delve further into this point, as the issues raised earlier already highlight more fundamental problems.''
\end{quote}
\subsubsection*{Response to R2.6}
This comment does not call for further action. In the General Discussion (subsection ``Encoding and retrieval interference"), we discuss that we believe that the difference in presentation rate and whether it was controlled by the participants or not led to different types of interference (encoding vs.\ retrieval).

### 2024-11-27 paper/revision2/ResponseToReviewers.tex

\section*{Reviewer \#2} 
\subsection*{Comment R2.1}
\begin{quote}
``I appreciate the authors' responses to my previous comments. However, despite the revisions, significant weaknesses remain, rendering in my opinion this study primarily a methodological contribution that adds little value to our understanding of interference effects. Additionally, the authors' strong claims often come through as overstatements that could result in serious misconceptions.
My overall assessment is as follows.

1. Interpretability of the SPR results: The results from the self-paced reading (SPR) task remain unclear. The only observable effect—the semantic interference—emerges well before the critical region of interest (specifically, at the distractor region) and persists without change beyond the critical region. As a result, any effect at the critical region is uninterpretable.

The authors acknowledge the interpretability issue with the SPR results in their response to my comment, and they now include the following statement: ``Given the pre-critical reading time differences, effects in the later regions, i.e., the critical and spill-over regions, cannot be attributed clearly to the stimuli in these regions and sentence processing mechanisms associated with them. Therefore, we refrain from further discussing effects occurring in the later regions." \textbf{However, this comes after more than 10 full pages (pp. 25-37) of discussion on the reaction times (RTs) in the critical and pre-critical regions, which would lead any reader to believe these findings are meaningful. If, as the authors now realize, the effect emerges much earlier, this entire discussion is misleading: the SPR results after the distractor region are simply uninterpretable.}''
\end{quote}

\subsubsection*{Response to R2.1}
We agree with the reviewer that the presentation and discussion of the SPR results was misleading and have revised it accordingly. We have revised Figure 3 to remove the focus on the critical region (see below). At the beginning of the Results section, we now anticipate our interpretation of the results and state that we believe that the distractors led to encoding interference. We also state that the results at the critical (and spill-over) region are potentially confounded because it is not possible to disentangle the long-lasting encoding effect from retrieval effects potentially arising in the critical region.

@@

\setcounter{figure}{2}

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C shows the reading times of the sentence focusing on the regions between the distractor and critical verb for all conditions.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled_zoom.jpg}
\end{sidewaysfigure}
\clearpage

The reviewer suggests that encoding interference confounded the reading times of all regions after the distractor and argues that this makes the results uninterpretable. The implication seems to be that the patterns observed at the pre-critical, critical, and spillover regions should not be reported.
 However, as we had already discussed in the results of the SPR results, the two published studies that used the present design \citep{vandyke07, mertzen} show clear indications that these regions of interest were affected by encoding interference as well. Thus, across the three studies (Van Dyke et al., Mertzen et al., and ours) one consistent pattern is that the observed effects could be driven at least partly by encoding interference. We feel that this consistent pattern across studies in the three regions of interest is worth reporting.

We discuss this on page \pageref{previous_encoding}:

@@

\begin{quote}
    \Paste{previous_encoding}
\end{quote}

To summarize, one valuable finding in our study is the replication of the pattern found in the previous studies, which is attributable to encoding interference. Ours is the only study among these three to prominently raise the possibility that the reading times of the critical region may be confounded by encoding interference--the reviewer would agree with this discussion. To ensure comparability to the previous studies, we consider it very important to report the reading times effects in the same regions which the previous studies reported, namely: the pre-critical, critical and spill-over region. This seems necessary to us despite the possible confound discussed above that the reviewer justifiably raises. We have revised the Results section to be more concise, starting on page \pageref{revised_SPR}:

### 2024-11-27 paper/revision2/ResponseToReviewers.tex

\begin{quote}
    \Paste{previous_encoding}
\end{quote}

To summarize, one valuable finding in our study is the replication of the pattern found in the previous studies, which is attributable to encoding interference. Ours is the only study among these three to prominently raise the possibility that the reading times of the critical region may be confounded by encoding interference--the reviewer would agree with this discussion. To ensure comparability to the previous studies, we consider it very important to report the reading times effects in the same regions which the previous studies reported, namely: the pre-critical, critical and spill-over region. This seems necessary to us despite the possible confound discussed above that the reviewer justifiably raises. We have revised the Results section to be more concise, starting on page \pageref{revised_SPR}:

\begin{quote}
    \Paste{revised_SPR}
\end{quote}

The main finding in our SPR study is that reading times were primarily affected by semantic encoding interference.  In order to fully understand the effect of semantic interference and to relate it to the previous literature, the reader needs to see the analyses at the pre-critical, critical, and spillover regions. We did not analyze the distractor itself and the immediately following words because these were not identical across conditions. Instead, we start the analysis at the pre-critical region, which was the same word across conditions (\textit{abends}, `in the evening', in the example item used for the x-axis labels in Figure 3 above). This pre-critical region serves as a proxy for investigating the effect of encoding interference caused by the distractor. This pre-critical word preceded the critical verb, thus the reading times at the pre-critical word are unlikely to show retrieval interference. To emphasize this we have renamed the former section ``Pre-critical effects" to ``Reading times difference prior to retrieval". We have revised the beginning of the Discussion on page \pageref{prior_to_retrieval} accordingly:

### 2024-11-27 paper/revision2/elsarticle-template-harv.tex
Updated via the Open Science Framework

We turn next to the event-related potentials experiment.

@@

There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \citep{gibsonthomas99}, syntactic local coherence effects \citep{taboretal04}, and the agreement attraction phenomenon across different languages \citep[e.g.,][]{wagersetal,lago_etal_2021,tucker2015representing}.  However, there exists  evidence inconsistent with these findings. Regarding grammaticality illusions in double-center embeddings, the grammaticality illusion does not occur in German and Dutch \citep{VSLK11,FrankEtAl2015}. Regarding syntactic local coherence effects, a large-sample study \citep{lcpaape2025} presents evidence against syntactic local coherence. Regarding agreement attraction, there seems to be at least one language with rich case marking (Czech) that does not show convincing evidence of number agreement attraction \citep{chromy2023number}. \Copy{gendisccaveat}{Thus, it seems that, cross-linguistically, syntax does generally play a central role in building incremental structure and in completing dependencies.}\label{gendisccaveat} \Copy{gendisccaveat2}{\textcolor{blue}{Moreover, it is entirely possible that a stronger syntactic manipulation than the one we used ends up showing a syntactic interference effect of the type that \citet{vandyke07} originally reported. Indeed, our own findings would need to be replicated, ideally by an independent research group, if we want to be sure that in the present design there is no evidence for syntactic interference. If such a replication attempt is carried out, it would also be useful to conduct an lab-based self-paced reading study to investigate our speculation earlier that the absence of syntactic interference in our online SPR study may have been due to participants adopting a good-enough processing strategy, leading to only semantic interference being detectable.}}\label{gendisccaveat2}

>>>>

\textcolor{blue}{In summary, the main finding of the Experiment 1 is that the reading times were primarily affected by semantic encoding interference.} We turn next to the event-related potentials experiment.

@@

There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \citep{gibsonthomas99}, syntactic local coherence effects \citep{taboretal04}, and the agreement attraction phenomenon across different languages \citep[e.g.,][]{wagersetal,lago_etal_2021,tucker2015representing}.  However, there exists  evidence inconsistent with these findings. Regarding grammaticality illusions in double-center embeddings, the grammaticality illusion does not occur in German and Dutch \citep{VSLK11,FrankEtAl2015}. Regarding syntactic local coherence effects, a large-sample study \citep{lcpaape2025} presents evidence against syntactic local coherence. Regarding agreement attraction, there seems to be at least one language with rich case marking (Czech) that does not show convincing evidence of number agreement attraction \citep{chromy2023number}. \Copy{gendisccaveat}{Thus, it seems that, cross-linguistically, syntax does generally play a central role in building incremental structure and in completing dependencies.}\label{gendisccaveat} \Copy{gendisccaveat2}{\textcolor{blue}{Moreover, it is entirely possible that a stronger syntactic manipulation than the one we used ends up showing a syntactic interference effect of the type that \citet{vandyke07} originally reported. Indeed, our own findings would need to be replicated, ideally by an independent research group, if we want to be sure that in the present design there is no evidence for syntactic interference.}}\label{gendisccaveat2}
